---
title: "Setting Prior Parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r}
source("./plot_priors.R")
```


## Data Scaling

In order to have prior distributions that are realistic for any provided dataset, we rescale the dataset such that the average value in the matrix is 100.

$$\tilde M = M\cdot \frac{100}{\bar M}$$
By default, we set hyperprior parameters such that the priors for each element of **$P$ and $E$ are centered at $10/\sqrt(N)$** so the prior of $\hat M = PE$ is also centered at 100. 

$$\mathbb{E}\left[(PE)_{kg}\right] = \mathbb{E}\left[\sum_nP_{kn}E_{ng}\right] = \sum_n\mathbb E\left[P_{kn}\right]\mathbb{E}\left[E_{ng}\right] = \sum_n \frac{10}{\sqrt{N}}\cdot \frac{10}{\sqrt{N}} = 100$$

This ensures that the scale of our priors is a reasonable starting place. *All results are translated back to the input scale.*

## Default Priors

### Gamma

```{r fig.height = 5, fig.width = 8}
dims <- list(G = 1, N = 5, K = 1)
plot_gamma_priors(dims, title = "Default Gamma Prior")
```

### Exponential

```{r fig.height = 5, fig.width = 8}
dims <- list(G = 1, N = 5, K = 1)
plot_exponential_priors(dims, title = "Default Exponential Prior")
```

### Truncated Normal

```{r fig.height = 5, fig.width = 8}
dims <- list(G = 1, N = 5, K = 1)
plot_truncnorm_priors(dims, title = "Default Truncated Normal Prior")
```

## Custom Hyperprior Parameters

Custom priors can be specified to `bayesNMF` with the `prior_parameters` argument. We recommend playing with the code below to ensure that the scale of the induced prior on $\hat M = PE$ is still reasonable. Be sure to include the correct dimensions of your data in the `dims` list.

### Gamma

$$
\begin{align*}
    P_{kn} &\sim Gamma(\alpha^P_{kn}, \beta^P_{kn})\\
    E_{ng} &\sim Gamma(\alpha^E_{ng}, \beta^E_{ng})\\\\
    \beta^P_{kn} &\sim Gamma(a^P_{kn}, b^P_{kn})\\
    \beta^E_{ng} &\sim Gamma(a^E_{ng}, b^E_{ng})\\
    \alpha^P_{kn} &\sim Gamma(c^P_{kn}, d^P_{kn})\\
    \alpha^E_{ng} &\sim Gamma(c^E_{ng}, d^E_{ng})\\
\end{align*}
$$

```{r fig.height = 5, fig.width = 8}
dims <- list(G = 10, N = 2, K = 10)
plot_gamma_priors(
    dims, title = "Custom Gamma Prior",
    prior_parameters = list(
        a_p = 20,
        b_p = 50,
        c_p = 10,
        d_p = 4
    )
)
```

### Exponential

$$
\begin{align*}
    P_{kn} &\sim Exp(\lambda = \lambda^P_{kn})\\
    E_{ng} &\sim Exp(\lambda = \lambda^E_{ng})\\\\
    \lambda^P_{kn} &\sim Gamma(a^P_{kn}, b^P_{kn})\\
    \lambda^E_{ng} &\sim Gamma(a^E_{ng}, b^E_{ng})
\end{align*}
$$
```{r fig.height = 5, fig.width = 8}
dims <- list(G = 100, N = 5, K = 96)
plot_exponential_priors(
    dims, title = "Custom Exponential Prior",
    prior_parameters = list(
        a_p = 200,
        b_p = 100,
        a_e = 1,
        b_e = 10
    )
)
```


### Truncated Normal

$$
\begin{align*}
    P_{kn} &\sim TruncNorm(\mu = \mu^P_{kn}, \sigma^2 = \sigma^{2P}_{kn}, a = 0, b = \infty)\\
    E_{ng} &\sim TruncNorm(\mu = \mu^E_{ng}, \sigma^2 = \sigma^{2E}_{ng}, a = 0, b = \infty)\\
    p(\sigma^2_k) &\propto 1/\sigma^2_k\\\\
    \mu_{kn}^P &\sim Normal(m_{kn}^P, s_{kn}^P)\\
    \sigma^{2P}_{kn} &\sim InvGamma(a_{kn}^P, b_{kn}^P)\\
    \mu_{ng}^E &\sim Normal(m_{ng}^E, s_{ng}^E)\\
    \sigma^{2E}_{ng} &\sim InvGamma(a_{ng}^E, b_{ng}^E)
\end{align*}
$$

Note that the parameter $\mu$ in a truncated normal is not the mean. For a truncated normal distribution with parameters $\mu$, $\sigma^2$ and bounded by $a$ and $b$, its mean is

$$
\mu + \sigma \cdot \frac{\varphi(\frac{a - \mu}{\sigma}) - \varphi(\frac{b - \mu}{\sigma})}{\Phi(\frac{b - \mu}{\sigma}) - \Phi(\frac{a - \mu}{\sigma})}
$$
where $\varphi$ is the standard normal pdf and $\Phi$ is the standard normal cdf.


```{r fig.height = 5, fig.width = 8}
dims <- list(G = 100, N = 5, K = 96)
plot_truncnorm_priors(
    dims, title = "Custom Truncated Normal Prior",
    prior_parameters = list(
        m_p = 100/dims$N,
        s_p = 0.0,
        m_e = 0.1,
        s_e = 1
    )
)
```
